---
title: "EEZs for Aquaculture on the West Coast"
author: "Lauren Puffer"
date: "2025-11-09"
format: html
editor: visual
embed-resources: true
page-loading: false
code-fold: show
execute: 
  warning: false
  message: false
toc: TRUE
theme: journal
---

# Introduction

# Load libraries

```{r}
library(tidyverse)
library(sf)
library(terra)
library(tmap)
library(patchwork)
library(here)
library(dplyr)
```

# Load data

SST files used:

-   `average_annual_sst_2008.tif`

-   `average_annual_sst_2009.tif`

-   `average_annual_sst_2010.tif`

-   `average_annual_sst_2011.tif`

-   `average_annual_sst_2012.tif`

Bathymetry files used:

-   depth.tif

## Exclusive Economic Zones

Exclusive Economic Zones files used:

-   wc_regions_clean.shp

```{r}
# Read in shapefile of EEZs
eez <- st_read(here("data", "wc_regions_clean.shp"))
```

## Sea surface temperature

```{r}
# Read in rasters of sea surface temerpature

sst_2008 <- rast(here("data", "average_annual_sst_2008.tif"))
sst_2009 <- rast(here("data", "average_annual_sst_2009.tif"))
sst_2010 <- rast(here("data", "average_annual_sst_2010.tif"))
sst_2011 <- rast(here("data", "average_annual_sst_2011.tif"))
sst_2012 <- rast(here("data", "average_annual_sst_2012.tif"))
```

## Bathymetry

```{r}
# read in raster for depth
bath <- rast(here("data", "depth.tif"))
```

## Oysters

-   sea surface temperature: 11-30°C

-   depth: 0-70 meters below sea level

## Stack rasters

```{r}
# Use c() to stack
sst_stack <- c(sst_2008, 
              sst_2009, 
              sst_2010,
              sst_2011, 
              sst_2012)
```

## Calculate mean

```{r}
# Get the mean of each of the cell values
sst_mean <- mean(sst_stack)

# Globally subtract 273.15
sst_c_mean <- sst_mean - 273.15   # subtract 273.15 from every cell
```

## Reproject bathymetry

```{r}
# Reproject to SST crs
bath_reproj <- project(bath, crs(sst_c_mean))

# Check that the rasters match
crs(bath_reproj) == crs(sst_c_mean)
```

Resample the bathymetry raster to the sea surface temperature.

```{r}
# Resample to SST
bath_aligned <- resample(bath_reproj, sst_c_mean)

# Check the extent
ext(bath_aligned) == ext(sst_c_mean)

# Check the resolution
res(bath_aligned) == res(sst_2008)
```

Create a matrix of oyster sea surface temperature threshold, and oyster depth range.

```{r}
# Matrix of SST rang
sst_osyter_matrix <- matrix(c(
  -Inf, 11, 0,
  11, 30, 1,
  30, Inf, 0),
ncol = 3,
byrow = TRUE)
  
# Matrix of depth range
depth_oyster_matrix <- matrix(c(
  -Inf, -70, 0,
  -70, 0, 1,
  0, Inf, 0),
  ncol = 3,
  byrow = TRUE
)
```

Reclassify the raster with the matrix.

```{r}
# Reclassify
sst_reclassify <- classify(sst_c_mean, sst_osyter_matrix)
bath_reclassify <- classify(bath_aligned, depth_oyster_matrix)
```

Multiply the rasters by eachother.

```{r}
# Multiply the reclassified rasters by eachother
multiplied <- sst_reclassify * bath_reclassify

# Change multiplied raster 0's to NAs
multiplied[multiplied == 0 ] <- NA

# Plot suitable cells
plot(multiplied)
```

## Make a mask of EEZ

```{r}
# Reproject EEZ to SST
eez_proj <- st_transform(eez, crs(multiplied))

# Check the CRS
crs(multiplied) == crs(eez_proj)
```

## Crop suitable cells to EEZ

```{r}
# Crop suitable cells to EEZ
multiplied_crop <- crop(multiplied, eez_proj)

# Plot the cropped 
plot(multiplied_crop)

# Rasterize the EEZ
eez_rast <- rasterize(eez_proj, multiplied_crop, field = "area_km2")
```

## Calculate zonal statistics 

```{r}
# Compue the area covered by each raster cell in km2
multiplied_crop_area <- cellSize(multiplied_crop, mask = TRUE, unit = "km")

# Calculate zonal statistics of cropped raster with EEZ raster
suitable_zonal <- zonal(multiplied_crop_area,eez_rast, fun = "sum", 
                        na.rm = TRUE) # remove NAs

# Use cbin to bring in
zones_bind <- cbind(eez_proj, suitable_zonal)

# Keep only the useful columns
zones_clean <- zones_bind |>
  select(rgn, area_km2, area_km2.1) |>
  rename(region = rgn, 
         total_area = area_km2, 
         area_suitable = area_km2.1)
```

## Map of suitable area per EEZ

```{r}
# Make a map of suitable area within Economic Zones
tm_shape(zones_clean)+
  tm_graticules()+
  tm_polygons(fill = "area_suitable")
```

# Function from suitable area workflow

We must repeat the process above in a generalizable function in order to be able to pass any species information through and get the same result.

To break down the process, our code needs to...

Take the inputs of 1) a raster with mean SST, 2) a raster of depth, 3) a shapefile, 4) max and min sst, 5) max and min sea surface temperature, and 5) species name.

With these data, the function must:

1.  Ensure that all CRSs are the same for EEZ, SST, and depth.
2.  Subtract 273.15 from the mean values of SST
3.  Crop depth to the extent of SST.
4.  Create matrix of max and min for SST and depth
5.  Apply matrix to SST and depth raster
6.  Multiply SST and depth suitable area rasters
7.  Crop suitable area raster to EEZ.
8.  Calculate zonal statistics for each EEZ based on suitable area.
9.  Make map of suitable area in each zone as a color gradient.

## EEZ Suitable area function

```{r}
# Mean SST pre calculated and 273.15 subtracted if in Kelvin
eez_suitable_area <- function(mean_sst, bath, eez, min_sst, max_sst, min_depth, max_depth, species_name){
  
  # Reproject bathymetry to match the CRS of SST
  if(crs(bath) == st_crs(mean_sst)){
  print("Coordinate reference systems match.")
}else{
  warning("Updating coordinate reference systems to match.")
  #transform data to match
  bath_reproj <- st_transform(bath, crs(mean_sst))
}
  
  # Crop bathymetry to match the extent of the SST raster
  bath_crop <- crop(bath_reproj, mean_sst)
  
  # Ensure that resolutions match
    # Reproject bathymetry to match the CRS of SST
  if(res(bath_crop) == res(mean_sst)){
  print("Spatial reoslutions match.")
  }else{
  warning("Updating resolutions to match.")
  # Resample data to match
  bath_aligned <- resample(bath_crop, y = sst_mean, method = "near")
  }
  
  #

}
```

Repeat process for ![](data:image/gif;base64,R0lGODlhAQABAIAAAP///wAAACH5BAEAAAAALAAAAAABAAEAAAICRAEAOw==)P.interruptus (California spiny lobster).
